---
title: "HAC Specifications: Version 3.2.2 (ICD-10-AM 13th Edition)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HAC Specifications: Version 3.2.2 (ICD-10-AM 13th Edition)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  echo       = FALSE,
  message    = FALSE,
  warning    = FALSE
)
library(icd10am.hacs)
library(knitr)

# Helper: return a kable of all numerator codes for a sub_category string match
hac_table <- function(sub_cat_pattern, role_filter = "numerator",
                      tbl = hac_table_a,
                      caption = NULL) {
  df <- subset(
    tbl,
    icd_edition == "icd10am_13" & hac_v3_2_2 == TRUE &
      role == role_filter &
      grepl(sub_cat_pattern, sub_category, fixed = TRUE)
  )
  if (nrow(df) == 0L) return(invisible(NULL))
  df <- df[order(df$sub_category, df$code_dot), c("sub_category", "code_dot", "description")]
  names(df) <- c("Sub-category", "Code", "Description")
  kable(df, row.names = FALSE, caption = caption)
}

# Helper: return a kable from a non-hac_table_a dataset
simple_table <- function(df, caption = NULL) {
  df <- df[order(df$sub_category, df$code_dot),
           c("sub_category", "code_dot", "description")]
  names(df) <- c("Sub-category", "Code", "Description")
  kable(df, row.names = FALSE, caption = caption)
}
```

## Background

A hospital-acquired complication (HAC) refers to a complication for which
clinical risk mitigation strategies may reduce (but not necessarily eliminate)
the risk of that complication occurring. The national list of 16 HACs was
developed by the Australian Commission on Safety and Quality in Health Care
(the Commission) through a comprehensive process that included reviews of the
literature, clinical engagement and testing of the concept with public and
private hospitals. The 16 complications can be broken down into 43 diagnoses
to allow for more detailed analysis. More information about the development of
the list is described on the Commission's website.

**Source:** <https://www.safetyandquality.gov.au/our-work/indicators/hospital-acquired-complications>

---

## Counting rules

HACs can be counted at the complication level (n=16) or the diagnosis level
(n=43), however, they should not be counted at the ICD-10-AM code level as
this may lead to double counting. For example, there are many codes used to
specify the diagnosis of 'Blood stream infection' --- a patient may have
multiple sepsis codes in one separation, e.g. unspecified sepsis, puerperal
sepsis and severe sepsis. This should be counted as one 'blood stream
infection', as these sepsis codes most likely refer to the same infection.

---

## Monitoring of HACs in different patient groups

Some patient groups have different rates of HACs. Monitoring of these patients
separately will allow safety and quality interventions to be targeted and
provide hospitals and clinicians with additional detail on the care they
provide. Each service will have different requirements and should consider their
casemix when deciding how to monitor HACs. Some groups that may be monitored
separately include, but are not limited to:

- Patients receiving mental health care
- Sub-acute patients
- Paediatric patients

---

## Intellectual property

The Commission owns all Intellectual Property Rights in the hospital-acquired
complications list subject to any third party Intellectual Property Rights that
may rest in the list. The Commission gives no warranty and makes no
representation as to the quality of the output of the list and users of the
list do so at their own risk.

The Commission recommends that the list is used as a part of a suite of
indicators to monitor safety and quality of care rather than as a stand-alone
measure.

The HACs list is provided free of charge for non-commercial purposes. If an
organisation would like to use the list for commercial purposes they should
request this in writing to the Commission by emailing
<indicators@safetyandquality.gov.au>.

---

## Provision of the list to third parties

To assist in version control, the Commission asks that users do not circulate
the HACs specifications outside of their organisation but to direct new users
to download a copy of the codes from the Commission's website.

---

## Revisions

The HACs list will be updated with each new edition of ICD-10-AM and released
on the Commission's website.

---

## Specifications for complications

This vignette shows diagnosis codes active under **HAC specification version
3.2.2** (September 2025) for **ICD-10-AM 13th edition** codes. All 619 active
numerator code rows across HACs 1--14, 15 and 16 are presented. HAC 5
(Unplanned ICU admission) is not represented by ICD-10-AM diagnosis codes ---
that complication is measured using an ICU admission data item not included in
the standard Admitted Patient Care collection.

For technical details of the classification algorithm see the function
documentation: `?identify_hacs`.

---

## HAC 1 -- Pressure injury

Pressure injuries (also known as pressure ulcers or bed sores) are areas of
damage to the skin and underlying tissue caused by unrelieved pressure. The HAC
covers Stages III and IV, unstageable, suspected deep tissue injury, and
unspecified/unclassified pressure injuries acquired during the admission.

```{r hac1-1}
hac_table("1.1", caption = "1.1 Stage III ulcer")
```

```{r hac1-2}
hac_table("1.2", caption = "1.2 Stage IV ulcer")
```

```{r hac1-3}
hac_table("1.3", caption = "1.3 Unspecified decubitus and pressure area")
```

```{r hac1-4}
hac_table("1.4", caption = "1.4 Unstageable pressure injury")
```

```{r hac1-5}
hac_table("1.5", caption = "1.5 Suspected deep tissue injury")
```

---

## HAC 2 -- Falls resulting in fracture or other intracranial injury

A fall-related HAC requires both a qualifying injury code (fracture or
intracranial injury) **and** a fall external-cause code (W01x, W03x, W04x,
W05x, W06x, W07x, W08x, W09x, W10x, W18) in a sequence that places the fall
after the injury diagnosis. Place-of-occurrence checking was removed in v3.x.

```{r hac2-1}
hac_table("2.1", caption = "2.1 Intracranial injury")
```

```{r hac2-2}
hac_table("2.2", caption = "2.2 Fractured neck of femur")
```

```{r hac2-3}
hac_table("2.3", caption = "2.3 Other fractures")
```

---

## HAC 3 -- Healthcare-associated infection

Healthcare-associated infections (HAIs) cover a broad range of infections
acquired during an admission. The sub-categories cover urinary tract
infections, surgical site infections, pneumonia, bloodstream infections,
catheter-related infections, multi-resistant organisms, prosthetic/implantable
device infections, gastrointestinal infections, and other high-impact
infections.

```{r hac3-1}
hac_table("3.1", caption = "3.1 Urinary tract infection")
```

```{r hac3-2}
hac_table("3.2", caption = "3.2 Surgical site infection")
```

```{r hac3-3}
hac_table("3.3", caption = "3.3 Pneumonia")
```

```{r hac3-4}
hac_table("3.4", caption = "3.4 Blood stream infection")
```

```{r hac3-5}
hac_table("3.5", caption = "3.5 Infections or inflammatory complications associated with peripheral/central venous catheters")
```

```{r hac3-6}
hac_table("3.6", caption = "3.6 Multi-resistant organism")
```

```{r hac3-7}
hac_table("3.7", caption = "3.7 Infection associated with prosthetics/implantable devices")
```

```{r hac3-8}
hac_table("3.8", caption = "3.8 Gastrointestinal infections")
```

```{r hac3-9}
hac_table("3.9", caption = "3.9 Other high impact infections")
```

---

## HAC 4 -- Surgical complications requiring unplanned return to theatre

Sub-categories 4.1 (haematoma), 4.3 (anastomotic leak), and 4.5 (other
complications) require additional algorithmic criteria beyond the diagnosis
code:

- **4.1 Haematoma** requires a haematoma-drainage or transfusion ACHI
  procedure code in the same episode.
- **4.3 Anastomotic leak** requires external-cause code Y83.2 (anastomosis
  procedure) to appear after the leak diagnosis.
- **4.2, 4.4, 4.5** are unconditional (diagnosis code alone is sufficient).

> **Note:** HAC 4 (unplanned return to theatre, sub-category 4.5 and related)
> is partially dependent on data items not currently collected in all
> Admitted Patient Care collections. Sub-categories 4.1--4.4 are implementable
> from standard NMDS data.

```{r hac4-1}
hac_table("4.1", caption = "4.1 Postoperative haemorrhage/haematoma requiring transfusion and/or return to theatre")
```

```{r hac4-2}
hac_table("4.2", caption = "4.2 Surgical wound dehiscence")
```

```{r hac4-3}
hac_table("4.3", caption = "4.3 Anastomotic leak")
```

```{r hac4-4}
hac_table("4.4", caption = "4.4 Vascular graft failure")
```

```{r hac4-5}
hac_table("4.5", caption = "4.5 Other surgical complications requiring unplanned return to theatre")
```

---

## HAC 5 -- Unplanned intensive care unit admission

Data items for this measure are not currently collected in the Admitted Patient
Care (APC) data collection. These data elements may be collected locally and
specifications for these data elements have been developed and are available on
METeOR. Accordingly, **no ICD-10-AM codes are associated with this HAC** and
it is not computed by `identify_hacs()`.

---

## HAC 6 -- Respiratory complications

Respiratory complications cover failure requiring mechanical ventilation,
aspiration pneumonia, and pulmonary oedema. Sub-category 6.1 requires an ACHI
mechanical ventilation procedure code in addition to the diagnosis code.

```{r hac6-1}
hac_table("6.1", caption = "6.1 Respiratory failure including acute respiratory distress syndromes requiring ventilation")
```

```{r hac6-2}
hac_table("6.2", caption = "6.2 Aspiration pneumonia")
```

```{r hac6-3}
hac_table("6.3", caption = "6.3 Pulmonary oedema")
```

---

## HAC 7 -- Venous thromboembolism

Venous thromboembolism includes pulmonary embolism and deep vein thrombosis
arising during an admitted episode.

```{r hac7-1}
hac_table("7.1", caption = "7.1 Pulmonary embolism")
```

```{r hac7-2}
hac_table("7.2", caption = "7.2 Deep vein thrombosis")
```

---

## HAC 8 -- Renal failure

Renal failure requiring haemodialysis or continuous veno-venous
haemodialysis. Sub-category 8.1 requires a dialysis ACHI procedure code in
the episode and excludes patients with pre-existing chronic kidney disease
Stage 4 (N18.4) or Stage 5 (N18.5).

```{r hac8-1}
hac_table("8.1", caption = "8.1 Renal failure requiring haemodialysis or continuous veno-venous haemodialysis")
```

---

## HAC 9 -- Gastrointestinal bleeding

```{r hac9-1}
hac_table("9.1", caption = "9.1 Gastrointestinal bleeding")
```

---

## HAC 10 -- Medication complications

Medication-related complications span four sub-categories. Sub-categories 10.1
and 10.4 require an external-cause code identifying the responsible drug to
appear after the complication diagnosis in the coded sequence. Sub-categories
10.2 and 10.5 are unconditional.

> Note: sub-category 10.3 (Opioid-related) is not included in version 3.2.2.

```{r hac10-1}
hac_table("10.1", caption = "10.1 Drug-related respiratory complications/depression")
```

```{r hac10-2}
hac_table("10.2", caption = "10.2 Haemorrhagic disorder due to circulating anticoagulants")
```

```{r hac10-4}
hac_table("10.4", caption = "10.4 Movement disorders due to psychotropic medication")
```

```{r hac10-5}
hac_table("10.5", caption = "10.5 Serious alteration to conscious state due to psychotropic medication")
```

---

## HAC 11 -- Delirium

```{r hac11-1}
hac_table("11.1", caption = "11.1 Delirium")
```

---

## HAC 12 -- Incontinence

```{r hac12-1}
hac_table("12.1", caption = "12.1 Urinary incontinence")
```

```{r hac12-2}
hac_table("12.2", caption = "12.2 Faecal incontinence")
```

---

## HAC 13 -- Endocrine complications

```{r hac13-1}
hac_table("13.1", caption = "13.1 Malnutrition")
```

```{r hac13-2}
hac_table("13.2", caption = "13.2 Hypoglycaemia")
```

---

## HAC 14 -- Cardiac complications

Sub-category 14.2 (Arrhythmias) has a split logic for bradycardia (R00.1):

- **cat14p2p1** -- R00.1 without a cardiac pacing procedure
- **cat14p2p2** -- R00.1 with a cardiac pacing procedure

Other arrhythmia codes in 14.2 are unconditional. Sub-categories 14.1, 14.3,
14.4, and 14.5 are also unconditional.

```{r hac14-1}
hac_table("14.1", caption = "14.1 Heart failure and pulmonary oedema")
```

```{r hac14-2}
hac_table("14.2", caption = "14.2 Arrhythmias")
```

```{r hac14-3}
hac_table("14.3", caption = "14.3 Cardiac arrest")
```

```{r hac14-4}
hac_table("14.4", caption = "14.4 Acute coronary syndrome including unstable angina, STEMI and NSTEMI")
```

```{r hac14-5}
hac_table("14.5", caption = "14.5 Infective endocarditis")
```

---

## HAC 15 -- Third and fourth degree perineal laceration during delivery

HAC 15 applies to vaginal births only. An episode is included if it has:

1. A numerator code (O70.2 or O70.3, Table 15a) **not** as the principal
   diagnosis
2. A vaginal birth outcome code (Table 15b) in the episode
3. **No** elective or emergency caesarean section ACHI code (Table 15c)
4. Admission mode is not a transfer from another facility
5. Care type is not palliative, geriatric evaluation/management, or
   maintenance (types 9, 10, 7.3)

The output flags are:

- `hac15p1` / `cat15p1p1` -- third degree laceration (O70.2)
- `hac15p2` / `cat15p1p2` -- fourth degree laceration (O70.3)

```{r hac15a}
b_data <- subset(hac_table_b,
                 icd_edition == "icd10am_13" & hac_v3_2_2 == TRUE)
b_out <- b_data[order(b_data$code_dot), c("sub_category", "code_dot", "description")]
names(b_out) <- c("Sub-category", "Code", "Description")
kable(b_out, row.names = FALSE,
      caption = "Table 15a: HAC 15 numerator codes (third/fourth degree perineal laceration)")
```

```{r hac15b}
c_data <- subset(hac_table_c,
                 icd_edition == "icd10am_13" & hac_v3_2_2 == TRUE)
c_out <- c_data[order(c_data$code_dot), c("sub_category", "code_dot", "description")]
names(c_out) <- c("Sub-category", "Code", "Description")
kable(c_out, row.names = FALSE,
      caption = "Table 15b: HAC 15 denominator codes (vaginal birth outcome)")
```

```{r hac15c}
d_data <- subset(hac_table_d,
                 icd_edition == "icd10am_13" & hac_v3_2_2 == TRUE)
d_out <- d_data[order(d_data$achi_code), c("sub_category", "achi_code", "description")]
names(d_out) <- c("Sub-category", "ACHI Code", "Description")
kable(d_out, row.names = FALSE,
      caption = "Table 15c: HAC 15 caesarean section exclusion codes (ACHI)")
```

---

## HAC 16 -- Neonatal birth trauma

HAC 16 applies to newborn episodes (care type 7) admitted from birth (not
transfers). Episodes with exclusion codes indicating extreme prematurity or
very low birth weight are excluded (Table 16b).

The output flags are:

- `hac16` (via `cat16p1`) -- neonatal birth trauma (16.1)
- `hac16` (via `cat16p2`) -- hypoxic ischaemic encephalopathy (16.2)

```{r hac16a}
e_data <- subset(hac_table_e,
                 icd_edition == "icd10am_13" & hac_v3_2_2 == TRUE)
e_out <- e_data[order(e_data$sub_category, e_data$code_dot),
                c("sub_category", "code_dot", "description")]
names(e_out) <- c("Sub-category", "Code", "Description")
kable(e_out, row.names = FALSE,
      caption = "Table 16a: HAC 16 numerator codes (neonatal birth trauma)")
```

```{r hac16b}
f_data <- subset(hac_table_f,
                 icd_edition == "icd10am_13" & hac_v3_2_2 == TRUE)
f_out <- f_data[order(f_data$code_dot), c("sub_category", "code_dot", "description")]
names(f_out) <- c("Sub-category", "Code", "Description")
kable(f_out, row.names = FALSE,
      caption = "Table 16b: HAC 16 exclusion codes (prematurity/low birth weight)")
```

---

## Summary: code counts by HAC and sub-category

```{r summary-table}
library(icd10am.hacs)

active_a <- subset(hac_table_a,
                   icd_edition == "icd10am_13" & hac_v3_2_2 == TRUE &
                     role == "numerator")

counts_a <- aggregate(code_dot ~ complication + sub_category,
                      data = active_a, FUN = length)
names(counts_a)[3] <- "n_codes"

# Add HAC 15 and 16
b16 <- data.frame(
  complication = "15. Third and fourth degree perineal laceration during delivery",
  sub_category = "15.1 Third and fourth degree perineal laceration during delivery",
  n_codes      = nrow(subset(hac_table_b, icd_edition == "icd10am_13" & hac_v3_2_2 == TRUE))
)
e16_counts <- aggregate(
  code_dot ~ sub_category,
  data = subset(hac_table_e, icd_edition == "icd10am_13" & hac_v3_2_2 == TRUE),
  FUN  = length
)
e16_counts$complication <- "16. Neonatal birth trauma"
names(e16_counts)[2] <- "n_codes"
e16_counts <- e16_counts[, c("complication", "sub_category", "n_codes")]

all_counts <- rbind(counts_a, b16, e16_counts)
all_counts <- all_counts[order(all_counts$complication, all_counts$sub_category), ]

names(all_counts) <- c("Complication", "Sub-category", "Codes (ICD-10-AM 13th ed)")
kable(all_counts, row.names = FALSE,
      caption = "Numerator code counts per sub-category, HAC v3.2.2 / ICD-10-AM 13th edition")
```
